Distribution-oblivious Programming with Dish

### 2019-11-20 pldi/draft.tex
Progress on the intermediate representation section, especially the part about the graph nodes

\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{soul}
\usepackage{xspace}
\usepackage{color}
\usepackage{xcolor}
\usepackage{upquote}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{cleveref}
\usepackage{wrapfig}

@@

% For comments
\newcommand{\eat}[1]{}
\newcommand{\TODO}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\todo}[1]{\hl{#1}\xspace}
\newcommand{\nv}[1]{[{\color{cyan}#1 --- Nikos}]}
\newcommand{\kk}[1]{[{\color{magenta}#1 --- KK}]}
\newcommand{\review}[1]{{\color{red}#1}}

@@

\section{Introduction}
\label{intro}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.49\textwidth]{./figs/dish_schematic.pdf}
\caption{
  \textbf{\sys high-level schematic.}
	\sys XYZ
}
\label{fig:schematic}
\end{figure}

@@

% \begin{figure*}[t]
% \centering
% \includegraphics[width=0.49\textwidth]{./figs/dish_overview.pdf}
% \caption{
%   \textbf{Applying \sys to $p_1$.}
% }
% \label{fig:example}
% \end{figure*}
%

@@

\subsection{Unix Composition, Informally}
\label{bg:pipelines}

% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
\unix provides several ways for succinct program composition.
Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
These programs produce output and consume input concurrently, and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
\begin{wrapfigure}{r}{0.06\textwidth}
  \vspace{-12pt}
  % \begin{center}
    % \includegraphics[width=0.48\textwidth]{gull}
    \includegraphics[width=0.06\textwidth]{./figs/dish_ex1.pdf}
  % \end{center}
  \vspace{-30pt}
\end{wrapfigure}
This interface exposes a stream of contiguous character lines.
A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
% generating errors or diagnostics to the standard 
% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
% TODO: \kk{We have to mention character vs line based parallelism somewhere, and that we choose to see lines as the data quantum}

Commands can be tuned by two mechanisms, namely command-line options and environment variables.
Options control the execution of a program---\eg changing the \ttt{sort}ing order.
The shell does not have any visibility into these options; 
  their parsing and evaluation is left entirely up to each command.
Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
They are organized in a map from names to values, both of which are strings, and can be evaluated at any point in the program---including as commands, flags, a shell feature termed parameter \emph{expansion}.

% TODO: Talk about security primitives?
Streams are a key feature of \unix, and go hand-in-hand with the file abstraction:
  files are simply persistent streams that can be manipulated by commands or pipelines.
\begin{wrapfigure}{r}{0.15\textwidth}
  \vspace{-15pt}
  % \begin{center}
    % \includegraphics[width=0.48\textwidth]{gull}
    \includegraphics[width=0.15\textwidth]{./figs/dish_ex3.pdf}
  % \end{center}
  \vspace{-25pt}
\end{wrapfigure}
Reading a file or a set of files generates a stream; 
  similarly, streams can be trivially redirected to the file system---by appending with the file identifier.
Files are named, in the sense that they correspond to a global identifier---for example \ttt{cat /x/y} returns the persistent stream associated with the identifier \ttt{/x/y}.
\unix exposes a few different types of files---\eg directories, symbolic links, named pipes (or FIFOs), and special character devices.
As a result, file-system identifiers part of a pipeline may resolve to in-memory structures or pseudo-devices (\eg \ttt{/dev/urandom}).

% At a high level, POSIX defines several file types:
% (-) regular files, storing lines of text or binary data,
% (d) directories, grouping together multiple other files,
% (l) symbolic links, pointing to other files,
% (p) named pipes or FIFOs, communication primitives that can be named and manipulated like files,
% (s) domain sockets or DSes, full-duplex communication primitives that support passing file descriptors, and
% (b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.

Special files are particularly interesting because they are resolved to OS structures.
One example is the special \ttt{/proc} filesystem (procfs) that presents information about processes and the operating system, offering a feature analogous to programmatic introspection:
  with \ttt{procfs} commands can inspect and alter their directory or the flags passed to them.
% Here are a few examples:
% (i) \ttt{/proc/PID/cmdline} contains the command that originally started the process,
% (ii) \ttt{/proc/modules} contains a list of the kernel modules currently loaded, and
% (iii) \ttt{/proc/PID/cwd} contains a symlink to the current working directory of the process.
Similarly, \unix exposes file-system handles for various ephemeral structures, including standard streams.
This bidirectional relationship between ephemeral streams and persistent names---the fact that streams can be named in the file-system and files can be converted to streams---is a key enabler of the runtime rebinding achieved during our transformations.

A small but powerful set of features is concerned with stream manipulation.
Some of these features are built into the shell;
  for example, given a file identifier, stream redirection operators can direct stdout or stderr to a file.
\begin{wrapfigure}{r}{0.15\textwidth}
  \vspace{-15pt}
  % \begin{center}
    % \includegraphics[width=0.48\textwidth]{gull}
    \includegraphics[width=0.15\textwidth]{./figs/dish_ex3.pdf}
  % \end{center}
  \vspace{-25pt}
\end{wrapfigure}
As described before, the file can be a FIFO dubbing as input to another command.
More interestingly, some commands allow splitting (resp. merging) a single stream (resp. many streams) into multiple streams (resp. a single stream);
  stream merging can be achieved by appending one stream after the end of another or by fusing (zipping) streams column-wise into a single (fat) stream.
% TODO: talk about \ttt{xargs}?

% Explicit Redirection, Implicit splitting and merging, This can be done through the file-system operators

The combination of file-system identifiers pointing to ephemeral streams and the stream redirection described above enable surprisingly expressive pipelines. %---for example:
The most interesting feature of these pipelines is the presence of cycles:
  a pipeline segment processes some input, part of which is extracting the next input to process.
One example use case, discussed in the evaluation section~\sx{eval} is a web crawler that feeds extracted URLs from the input stream back to the start of the pipeline.
% These streams are created by the C library (glibc) at the start of program execution,
% and new streams can be created to connect to files, sockets, pipes,

Apart from pipes, the \unix shell provides several other forms of program composition such as sequential (\ttt{;}) and parallel (\ttt{\&}) composition operators.
Command substitution, expressed as \ttt{\$($c$)}, replaces the expression with the results of executing $c$.
Code blocks, expressed as \ttt{\{$c_1; c_2$\}}, can redirect streams for the entire group of $c_1$ and $c_2$;
  this is analogous to an anonymous function, which contains many statements and can possibly be assigned to a variable that can later be invoked with parameters.
Subshells, enclosed in \ttt{($c_1; c_2$)}, are similar to code blocks but spawn a child shell, thus avoiding the pollution of their environment variable context. % but incurring higher overhead.

% \begin{lstlisting}[
%   language=es,
%   mathescape,
%   float=t,
%   belowskip=-5mm,
%   label=core-pl,
%   upquote=true,
%   caption={
%     \textbf{The core language of the POSIX shell.}
%     The language used by ; file identifiers and expansion are not shown.
%   }]
% $p$ ::= $s\in{}String$ | $n\in{}Number$ | $b\in{}Bool$ | $\emptyset$
% $v$ ::= $p$ | ($x$,$\ldots$) => {e} | {$s$:$v$,$\ldots$} | [$v$,$\ldots$]
% $e$ ::= x | $v$ | ($x$ = $e$) $e$ | $e$($e$) | $e$[$e$] = $e$
% \end{lstlisting}

% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
%  cat /proc/ >                            (($$p_4$$))
% \end{lstlisting}
% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
%  ls -l | wc -l                           (($$p_2$$))
% \end{lstlisting}
% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
%  cat * | wc -l > out.txt                 (($$p_3$$))
% \end{lstlisting}
% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
%  cat /proc/ >                            (($$p_5$$))
% \end{lstlisting}
% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
% { echo 1; echo 2; } > out.txt
% \end{lstlisting}

\begin{figure}[t]
\centering
\includegraphics[width=0.49\textwidth]{./figs/dish_overview.pdf}
\caption{
  \textbf{\sys transformation overview.}
	\sys XYZ
}
\label{fig:overview}
\end{figure}

% \heading{Generation}
% Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
% The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams

\subsection{Overview}

@@

@@

@@

\kk{Text point: Problem: Not all parts of a shell script are
  distributable, so we cannot arbitrarily distribute}

@@

\subsection{Intermediate Representation/Dataflow graph}

Mention its constructs, and its semantics, and why it is a bit
different that standard dataflow graphs (where usually incoming edges
are either consumed with some arbitrary interleaving, or pair by
pair).

\kk{Text point: The main shell abstraction is files, which are batches
  or streams of characters (or lines) where the order matters. So we
  want a representation that captures order of inputs and outputs (and
  ours does because its node in the graph has a sequence of inputs.}

Explain that the intermediate representation represents workers that
communicate through files, and that it captures parallelism both in
the operator level, as well as data parallelism.

@@

\kk{Maybe this should be merged with the one above, especially if it
  talks about how distribution happens on the intermediate
  representation.}

\subsection{Scaling up the distributed computation/Vertically expanding the graph}

@@

@@

\begin{itemize}
\item Fault Tolerance
\item Distributed Planning: The planner at the moment is naive, and
  there is a huge literature on operator placement of distributed
  dataflow graphs that can be used to improve it.
\item Support of shell constructs: At the moment we have been
  conservative in the shell subsets that we handle, so that we don't
  introduce any unsafe parallelism. The point of this work was not to
  be able to handle a complete distributable subset of the shell, but
  rather a significantly big part of it, to demonstrate that there can
  be performance benefit from it.
\item Support distributing different classes: At the moment we are
  only dealing with stateless and pure commands, that is the ones that
  are most popular but also can offer the biggest benefits from
  parallelizing them. However, it would be interesting to explore the
  distributability of the file system commands, by using a distributed
  file system. THis would also require considerations about
  replication, consistency, and failures.
\end{itemize}

\section{Related Work}
\label{related}

\section{Conclusion}
\label{discussion}

>>>>

\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{soul}
\usepackage{xspace}
\usepackage{color}
\usepackage{xcolor}
\usepackage{upquote}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{cleveref}

@@

% For comments
\newcommand{\eat}[1]{}
\newcommand{\TODO}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\todo}[1]{\hl{#1}\xspace}
\newcommand{\nv}[1]{[{\color{cyan}#1 --- Nikos}]}
\newcommand{\kk}[1]{[{\color{magenta}#1 --- kk}]}
\newcommand{\review}[1]{{\color{red}#1}}

@@

\section{Introduction}

@@

@@

\begin{figure*}[t]
\centering
\includegraphics[width=0.49\textwidth]{./figs/dish_overview.pdf}
\caption{
  \textbf{Applying \sys to $p_1$.}
}
\label{fig:example}
\end{figure*}

@@

\subsection{Unix Pipelines, Informally}
\label{bg:pipelines}

Pipelines are a mechanism for program composition.
Individual programs, specified as commands, conform to a common but general interface. 
For now, this interface allows a combination of the following:
  accepting input from the standard input stream (stdin),
  generating output to the standard output stream (stdout), and
  generating errors or diagnostics to the standard error stream (stderr)
Standard streams are streams of characters with a few special characters---for example, EOF is a special end-of-file character signaling the end of the input stream.
% For Unix commands, the newline character is special; as such, Unix streams can be seen as lazy lists of strings---lines of text separated by the newline character.

Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
Pipeline chaining goes hand-in-hand with the Unix toolbox philosophy: 
  programs do one thing well, they work together, and they manipulate text streams as their universal interface.
The next few paragraphs describe the details.

\heading{Basics}
The standard shell syntax for pipelines a list of commands separated by vertical bars (``pipes'', in Unix terminology).

that are not interpreted by the shell, but are local to each command.

Additional details augment this simple interface.
Commands can be specified with options arguments

\begin{figure}[t]
\centering
\includegraphics[width=0.45\textwidth]{./figs/dish_ex1.pdf}
\label{fig:example}
\end{figure}

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 ls -l | wc -l                           (($$p_2$$))
\end{lstlisting}

\heading{Generation}
A central concept in pipelines---and Unix, in general---is the file system, which allows interfacing with persistent data.
There are many commands that take as input a file identifier 
Reading a file or a set of files generates a stream; 
  similarly, streams can be trivially redirected to the file system---by appending with the file identifier.

Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 cat * | wc -l > out.txt                 (($$p_3$$))
\end{lstlisting}

\heading{Identifiers}

\kk{We have to mention character vs line based parallelism somewhere,
  and that we choose to see lines as the data quantum}

The file-system is a common abstraction layer used for many other operations than just serving files (this will be become clearer when talking about file-system commands in \S\ref{bg:cmd}).
as such, different identifiers many not resolve to actual files, but rather in-memory or OS-internal structures.

At a high level, POSIX defines several file types:
(-) regular files, storing lines of text or binary data,
(d) directories, grouping together multiple other files,
(l) symbolic links, pointing to other files,
(p) named pipes or FIFOs, communication primitives that can be named and manipulated like files,
(s) domain sockets or DSes, full-duplex communication primitives that support passing file descriptors, and
(b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.

Most of these identifiers are manipulated by file-system commands---\eg \ttt{mkdir} for directories and \ttt{mkfifo} for FIFOs.
FIFOs and DSes look like files but are streams.

Special files are particularly interesting because they are resolved to OS structures.
One set is pseudo-devices---device files that simply;
For example /dev/urandom returns a stream of random characters and /dev/null accepts and discards all input written to it.
Another set is proc filesystem (procfs), a special filesystem in Unix-like operating systems that presents information about processes and other system information.
For example, /proc/PID/cmdline contains command that originally started the process and /proc/PID/cwd contains a symlink to the current working directory of the process.

sysfs is a pseudo file system provided by the Linux kernel that exports information about various kernel subsystems, hardware devices, and

/modules, one of the most important files in /proc, containing a list of the kernel modules currently loaded . It gives some indication (not always entirely correct) of dependencies.
/cpuinfo

symbolic links, FIFO specials, 
device files
block special, character special, 
and domain sockets.

There are 3 standard streams (stdin, stdout, stderr) created automatically for most Unix programs.
These correspond to \ttt{/dev/fd/0}.
% These streams are created by the C library (glibc) at the start of program execution,
and new streams can be created to connect to files, sockets, pipes,

Unix already specifies a few classes of files:

Here are a few special files:
  \ttt{proc} called

Technically, anonymous pipes show up as file identifiers in a hidden filesystem.
Similarly, open files and other ephemeral identifiers

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 cat /proc/ >                            (($$p_4$$))
\end{lstlisting}

\heading{Stream Manipulation}
A small but powerful set of primitives are concerned with stream manipulation.
First, as described above, several commands can be redirected
Second,

\ttt{xargs}
\ttt{tee}

In combination, these constructs are powerful enough to introduce circular streams. % or stream cycles?

Explicit Redirection
Implicit splitting and merging 
This can be done through the file-system operators

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 cat /proc/ >                            (($$p_5$$))
\end{lstlisting}

\heading{Composition}
anonymous pipes

Composition

The shell provides several ways of grouping commands and manipulating their
overall input and output channels

{ echo 1; echo 2; } > out.txt

@@

\section{Intermediate Representation/Dataflow graph}

In order to effectively apply optimizations and distribute a shell
script, it first has to be translated to a manipulable
representation. As mentioned in \ref{}, the two main abstractions of
the shell, are files that contain data, and commands that communicate
through these files. We propose a slight variation of the standard
dataflow graph, that is commonly found in stream processing systems,
as the representation of a distributed shell computation. The dataflow
graph nodes represent commands, and the edges represent files.

\kk{Before that I have to mention that the basic data type $D$ is a
  line of characters.}

\paragraph{Nodes - Commands}

Nodes $f$ of the graph represent functions from one (possibly empty)
input stream to an output stream $f : D^* \rightarrow D^*$.  This
representation captures the majority of shell commands. The nodes $f$
are monotonic, meaning that they cannot retract their output. More
formally, if $f(x) = y$, then for any $x'$, there exists $y'$ such
that $f(x.x') = y.y'$, where $.$ is standard concatenation and will be
omitted from now on. An example of a node is the following command
\kk{Does it need ref? Also maybe this example is too trivial and can
  be removed.}:

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 grep ``foo''
\end{lstlisting}

\noindent
This command takes a stream of lines from its standard input and
returns a stream of lines on its standard output. An important
observation though is that commands in the shell might read their
input stream from a sequence of different sources. An illustrative
example is the command \kk{Inline listing}cat which reads its input
from a combination of files and its standard input.

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 cat x - y
\end{lstlisting}

\kk{Should I reference that this example was takes from the cat
  manpage?} The example above reads file \texttt{x} until it
encounters the end of file, then reads from \texttt{stdin}, and then
reads from \texttt{y}. In order to handle this particularity, nodes in
the dataflow graph can have many incoming edges, that are clearly
ordered in a sequence. Note that this comes in contrast with standard
dataflow graph representations where multiple incoming edges either
mean zipping the input to pairs, or arbitraliy interleaving the
different inputs.

\kk{A note about nodes and commands that can write and read from more
  files that the ones in their streams}

Note that nodes can read from more files than just the files of their
stream (or respectively write to more files that their output stream),
however these files are static (e.g. used for configuration or logging
\kk{Would an example here help}) and do not represent streams of data,
and therefore are not considered part of the dataflow graph. Because
of the assumption in section Front End \ref{}, that each file exists
only once in the dataflow graph, we can safely assume that reading and
writing to these static files does not interfere with the distributed
implementation, and doesn't alter the behaviour of the program.

\paragraph{Edges - Files}

\kk{Maybe edges has to be moved before nodes, so that we can define
  the line data type}

Edges in the dataflow graph represent files, the basic data
abstraction of the shell. They are used as communication channels
between nodes in the graph, and as the input or output of the whole
graph. Edges are represented as possibly unbounded streams of type
$D*$, where $D$ represents a line, that is a sequence of characters
followed by \verb|\n| \kk{or EOF?}.

\kk{The following paragraph falls in the trap of just explaining
  without motivating}

Edges could either refer to a specific named file in the file system,
or just be FIFO pipes that used for interprocess communication \kk{Are
  there any other file types, such as URLs, ...?}. Edges that refer to
real files, contain their exact number of lines, while edges that are
fifo pipes can have an upper bound on the number of lines that they
will transfer through them, thus being bounded in size.

\kk{Depending on which of the two paragraphs goes last (nodes/edges),
  I have to explain there that the last incoming edge to a node must
  be unbounded (if it doesn't point to a resource).}

\kk{Text point: The main shell abstraction is files, which are
  sequences of lines, where the order matters. So we want a
  representation that captures order of inputs and outputs (and ours
  does because its node in the graph has a sequence of inputs.}

\paragraph{Merge and Split}

\kk{Mergers and splitters, and how they are implemented using shell constructs (cat and head)}

As we described above, a node could have many incoming edges

\kk{Section notes:}

\kk{Files have an associated number of lines}

\kk{Explain how the assumptions made in the front end give as correct
  distribution}

@@

\kk{Maybe this becomes a subsection either in the intermediate
  representation or in the planner.}

@@

\kk{Text point: Problem: Not all parts of a shell script are
  distributable, so we cannot arbitrarily distribute. Here we have to
  make our assumptions clear}

@@

\kk{The same file exists once in the substree. This is a conservative
  assumption that ensures that no two commands write or read to the
  same file concurrently.}

\kk{How can we make sure that nodes in the IR are deterministic. Do we
  make some assumption that enforces this?}

@@

\kk{Mention that different nodes can be safely executed independently,
  and that this exposes operator level parallelism}

\subsection{Scaling up the distributed computation/Vertically expanding the graph/Graph Transformations/ Equations}

@@

\kk{Can we have a semi-formal argument here?}

\kk{Explain how our graph transformations expose data parallelism as
  operator level parallelism.}

\kk{Different transformations to explain: Stateless, Pure, File
  splits, XARGS. Mention as much as possible formally using the
  definition of a node.}

The following equation holds for any stateless command $f$.

\[
f(x.x') = f(x).f(x')
\]

This is the basis of all optimizations that the graph does, and allows
for correct data parallelism.

@@

\begin{itemize}
\item Fault Tolerance
\item Distributed Planning: The planner at the moment is naive, and
  there is a huge literature on operator placement of distributed
  dataflow graphs that can be used to improve it.
\item Support of shell constructs: At the moment we have been
  conservative in the shell subsets that we handle, so that we don't
  introduce any unsafe parallelism. The point of this work was not to
  be able to handle a complete distributable subset of the shell, but
  rather a significantly big part of it, to demonstrate that there can
  be performance benefit from it.
\item Support distributing different classes: At the moment we are
  only dealing with stateless and pure commands, that is the ones that
  are most popular but also can offer the biggest benefits from
  parallelizing them. However, it would be interesting to explore the
  distributability of the file system commands, by using a distributed
  file system. THis would also require considerations about
  replication, consistency, and failures.
\item No cycles
\end{itemize}

\section{Related Work}
